---
title: "Homework 1 solution"
author: Joshua Loftus
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 4)
library(tidyverse)
library(gapminder)
```

# Random variables

## R's four functions for each distributions

### Question 1: Normal random variables

```{r}
mu <- sqrt(2)
sigma2 <- 9
# Expected value E[Z] = mu
mu
# Variance Var(Z) = sigma2
sigma2
# Distribution function: Prob(Z = 1)
# [note: for continuous random variables this is only approximate]
dnorm(1, mean = mu, sd = sqrt(sigma2))
# Cumulative distribution function: Prob(Z <= 1.5)
pnorm(1.5, mean = mu, sd = sqrt(sigma2))
# Upper tail instead: Prob(Z > 1.5)
1- pnorm(1.5, mean = mu, sd = sqrt(sigma2))
# Quantile: what is the specific value z
# such that P(Z <= z) = .95?
qnorm(.95, mean = mu, sd = sqrt(sigma2))
# Random sampling: generate a sample of 8 observations of Z
rnorm(8, mean = mu, sd = sqrt(sigma2))
```

## Plotting distribution functions

### Normal cumulative distribution function

Plotting the normal cdf:

```{r}
# copy and paste the code above, change dnorm to pnorm -- that's it!
range <- data.frame(x = c(-7, 10))
ggplot(range, aes(x)) + 
  stat_function(fun = pnorm, args = list(mean = mu, sd = sqrt(sigma2))) +
  theme_minimal()
```

### Question 2

Create a histogram plot like the one above, but with a larger sample size.

```{r}
# Copy and paste the code above
# Change sample_size to 1000 and nbins to 40
sample_size <- 1000
nbins <- 40
Z_sample <- data.frame(z = rnorm(sample_size, mean = mu, sd = sqrt(sigma2)))
ggplot(Z_sample, aes(z)) + geom_histogram(aes(y = ..density..), bins = nbins) +
  stat_function(fun = dnorm, args = list(mean = mu, sd = sqrt(sigma2))) +
  theme_minimal()
```

# Real data

## GDP and life expectancy in gapminder

Let's use data from the year 2002

```{r}
world <- filter(gapminder, year == 2002,
                continent != "Oceania")
```

### Problem 3

Histogram of `lifeExp` for the world.

```{r}
# paste here, change gdpPercap to lifeExp
ggplot(world, aes(lifeExp)) + 
  geom_histogram(bins = 15) +
  theme_minimal()
```

Density plots of `lifeExp` for each continent.

```{r}
# paste here, same as above
ggplot(world, aes(lifeExp)) + 
  geom_density() +
  facet_wrap(~continent, scales = "free") +
  theme_minimal()
```
