---
title: "Homework 1 (due Wednesday, April 4th)"
author: I agree to abide by the Stern Code of Conduct - CHANGE THIS TO YOUR NAME
output: html_document
---
---
title: "Homework 1 (due Thursday, April 5th)"
author: I agree to abide by the Stern Code of Conduct - CHANGE THIS TO YOUR NAME
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 4)
library(tidyverse)
library(gapminder)
# If you haven't done it yet, make sure to run
# install.packages("tidyverse") and
# install.packages("gapminder") in the console (below)
# before doing anything else
# If it asks if you want to compile packages from source
# choose "no" or you may get errors
```

# Random variables

## R's four functions for each distributions

`R` has four important functions for each of many different kinds of random variables. Run `?distributions` in the console to see which kinds of random variables are supported.

### Example: Binomial random variables

- This section is an example for you to follow to answer the questions in the next section.
- Run `?dbinom` to see the documentation for the four functions for the Binomial distribution.
- Notation: we say the random variable X has Binomial distribution Bin(n, p) with n trials and success probability p. 

```{r}
n <- 40
p <- 2/3
# Expected value E[X] = n*p
n*p
# Variance Var(X) = n*p*(1-p)
n*p*(1-p)
# Distribution function: Prob(X = 25)
dbinom(25, size = n, prob = p)
# Cumulative distribution function: Prob(X <= 21)
pbinom(21, size = n, prob = p)
# Upper tail instead: Prob(X > 21)
1 - pbinom(21, size = n, prob = p)
# Quantile: what is the specific value x
# such that P(X <= x) = .95?
qbinom(.95, size = n, prob = p)
# Random sampling: generate a sample of 8 observations of X
rbinom(8, size = n, prob = p)
```

### Question 1: Normal random variables

- Enter your code in the code chunk below to fill out this section similar to the Binomial example.
- Type `?dnorm` in the console to see the documentation for the functions for a normal random variable. 
- Notation: we say that Z has a normal distribution N(mu, sigma2) with mean mu and variance sigma2.
- Notation: `R` always uses the letter x to refer to values of the random variable, but we'll use z for this one.
- Warning: the `R` functions for normal use standard deviation, sd, instead of variance.
- Hint: use `sd = sqrt(sigma2)`.

```{r}
mu <- sqrt(2)
sigma2 <- 9
# Expected value E[Z] = mu
mu
# Variance Var(Z) = sigma2
sigma2
# Distribution function: Prob(Z = 1)
# [note: for continuous random variables this is only approximate]

# Cumulative distribution function: Prob(Z <= 1.5)

# Upper tail instead: Prob(Z > 1.5)

# Quantile: what is the specific value z
# such that P(Z <= z) = .95?

# Random sampling: generate a sample of 8 observations of Z

```

## Plotting distribution functions

### Normal probability distribution function

```{r}
range <- data.frame(x = c(-7, 10))
```

Plotting the normal pdf:

```{r}
# Copy this
ggplot(range, aes(x)) + 
  stat_function(fun = dnorm, args = list(mean = mu, sd = sqrt(sigma2))) +
  theme_minimal()
```


### Normal cumulative distribution function

Plotting the normal cdf:

```{r}
# copy and paste the code above, change dnorm to pnorm -- that's it!
```

### Histogram of a normal sample

```{r}
sample_size <- 100
nbins <- 15
Z_sample <- data.frame(z = rnorm(sample_size, mean = mu, sd = sqrt(sigma2)))
ggplot(Z_sample, aes(z)) + geom_histogram(aes(y = ..density..), bins = nbins) +
  stat_function(fun = dnorm, args = list(mean = mu, sd = sqrt(sigma2))) +
  theme_minimal()
```

### Question 2

Create a histogram plot like the one above, but with a larger sample size.

```{r}
# Copy and paste the code above
# Change sample_size to 1000 and nbins to 40
```

# Real data

## GDP and life expectancy in gapminder

Let's use data from the year 2002

```{r}
world <- filter(gapminder, year == 2002,
                continent != "Oceania")
```

### GDP per capita histogram

```{r}
# Copy this
ggplot(world, aes(gdpPercap)) + 
  geom_histogram(bins = 15) +
  theme_minimal()
```

### GDP per capita density

We can also easily create a smooth version of the histogram using `geom_density`, and have different panels for each continent using `facet_wrap`.

```{r}
# Copy this
ggplot(world, aes(gdpPercap)) + 
  geom_density() +
  facet_wrap(~continent, scales = "free") +
  theme_minimal()
```

### Problem 3

Histogram of `lifeExp` for the world.

```{r}
# paste here, change gdpPercap to lifeExp
```

Density plots of `lifeExp` for each continent.

```{r}
# paste here, same as above
```