---
title: "LMlab"
author: "Your name here!"
date: "4/23/2018"
output: html_document
---
---
title: "LMlab"
author: "Your name here!"
date: "4/23/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
######################################################################
############# Run the following line once then delete it #############
######################################################################
install.packages("ggfortify")
######################################################################
########### Delete the previous line after running it once ###########
######################################################################
library(tidyverse)
library(ggfortify)
library(fivethirtyeight)
library(gapminder)
theme_set(theme_minimal())
```

## First load the data and look at it

### First few rows of data

Use the `head()` function to show the first few rows of `candy_rankings`

```{r}

```

### Plot points we plan to predict

Store the plot in a new variable so we can reuse it several times:

```{r}
candy_plot <- ggplot(candy_rankings, aes(sugarpercent, winpercent)) +
  geom_point() 
```

Look at the plot the first time:

```{r}
candy_plot
```

## Fit a linear model

### The `lm` function

We use the function `lm(outcome_var ~ preditor_var, data = name_of_data)`

- Change `outcome_var` to the name of the variable you are trying to predict
- Change `predictor_var` to the name of the variable you are using as a predictor
- Change `name_of_data` to the name of the data set that has your variables
- Store the output of this function in a new variable so we can reuse it multiple times:

```{r}
#candy_model <- lm(change this)
```

Look at the basic information about the fitted model, including the estimated coefficients. Type `candy_model` below

```{r}

```
- Can also use `coef()` function
- Check: `coef(candy_model)[1]` and `coef(candy_model)[2]`

Add the regression line to our previous plot. Add `stat_smooth(method = "lm", se = F)` to the `candy_plot` below

```{r}
#candy_plot + change this
```

### Inference for the model

Use the `summary()` function on `candy_model` to get (1) p-values for the coefficients and (2) the R squared measure of goodness of fit

```{r}

```

Use `confint()` on the `candy_model` to get confidence intervals for the coefficients

```{r}

```

### Diagnostic plots to check the model

Use `autoplot()` function (from the `ggfortify` package) on `candy_model` to see multiple diagnostic plots for the fitted model:

```{r}

```

Create a scatterplot plot with `sugarpercent` (the predictor variable) on the horizontal axis and predicted outcomes on the vertical axis (hint: use `predict(candy_model)`)

```{r}

```

