---
title: "Scatterplots and regression lines"
author: "Joshua Loftus"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggthemes)
library(ggrepel)
library(fivethirtyeight)
library(gapminder)
set.seed(1)
```

#### Outline

- Creating scatterplots
- Adding regression lines
- Adding smooth curves
- Interpretation

## Scatterplots

### This doesn't show all the data! 

```{r}
mpg2008 <- filter(mpg, year == "2008")
mpgplot <- ggplot(mpg2008, aes(cty, hwy)) +
    theme_tufte() + 
    ggtitle("Miles per gallon") +
    geom_text_repel(data = subset(mpg2008, cty >= 26 | cty == 9),
                      mapping = aes(label = model))
mpgplot + geom_point()
```

### Alpha: transparency. Now overlapping points are darker

```{r}
mpgplot + geom_point(alpha = .5)
```

### "Jittering" the points moves them slightly so they don't overlap

```{r}
mpgplot + geom_point() + geom_jitter()
```

### "Rug" plots are like 1-dimensional histograms for each axis

```{r}
movies <- bechdel[complete.cases(bechdel),]
movie_plot <- ggplot(movies, aes(budget_2013, intgross_2013)) +
  geom_rug(size = .1, alpha = .2) +
  ggtitle("Budget vs gross earnings") +
  theme_tufte()
movie_plot + geom_point(alpha = .2)
```

### Transforming variables / changing axis scales

```{r}
log_movie_plot <- movie_plot +
  geom_point(alpha = .2, aes(color = binary)) +
  scale_x_log10() + scale_y_log10()
log_movie_plot
```

### Preview of more cool things

```{r}
log_movie_plot + geom_smooth(method = "loess", se = FALSE, color = "black",
                             aes(linetype = binary),
                             data = subset(movies, budget_2013 > 5e5)) +
  geom_abline(intercept = 0, slope = 1)
```

## Regression lines

### We'll go into lots of detail on linear models soon

```{r}
gm2007 <- gapminder %>% filter(year == 2007)
model <- lm(lifeExp ~ gdpPercap, data = gm2007)
gm2007$predicted <- predict(model)
```

### But first, let's visualize them

```{r}
gm_plot <- ggplot(gm2007, aes(gdpPercap, lifeExp)) + theme_tufte() +
  ggtitle("GDP and life expectancy") +
  geom_text_repel(data = subset(gm2007, gdpPercap > 49000 | lifeExp < 40),
                      mapping = aes(label = country)) +
  geom_point(alpha = .5) 
gm_plot + geom_smooth(method = "lm", se = FALSE)
```

### Errors of the model

```{r}
gm_plot + geom_smooth(method = "lm", se = FALSE) +
  geom_segment(alpha = .5,
               aes(x=gdpPercap, xend=gdpPercap, 
                   y=lifeExp, yend=predicted),
               data = subset(gm2007, gdpPercap > 12000))
```
