---
title: "Lecture 27 - Prediction and errors"
author: "Joshua Loftus"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggthemes)
library(ggrepel)
library(fivethirtyeight)
library(gapminder)
options(scipen=999)
set.seed(1)
```

#### Outline

- Computing coefficients in R
- Predictions and errors
- Extrapolation

### Computing coefficients

- Last time we learned how to calculate sample estimates $\hat \beta_0$ and $\hat \beta_1$ of the regression coefficients
- Use sample standard deviations and correlation to get the slope:
$$
\hat \beta_1 =  r \cdot \frac{s_y}{s_x}
$$
- Then use sample means and slope to get the intercept:
$$
\hat \beta_0 = \bar y - \hat \beta_1 \bar x
$$

- Calculating two variable summary in `R`:

```{r}
gm2007 <- filter(gapminder, year == "2007")
gm2007 %>% summarize(
  meanx = mean(gdpPercap),
  meany = mean(lifeExp),
  sdx = sd(gdpPercap),
  sdy = sd(lifeExp),
  r = cor(lifeExp, gdpPercap)
)
```

- Calculating regression coefficients from two variable summary

```{r}
beta1 <- cor(gm2007$lifeExp, gm2007$gdpPercap) * sd(gm2007$lifeExp) / sd(gm2007$gdpPercap)
beta0 <- mean(gm2007$lifeExp) - beta1 * mean(gm2007$gdpPercap)
c(beta0, beta1)
```

- Compare to coefficients calculated by `lm`

```{r}
gm_model <- lm(lifeExp ~ gdpPercap, data = gm2007)
gm_model
```

## Prediction and errors

### For points in the data

- Each data point has an actual outcome $y_i$, but we can also predict the outcome using the regression model
- Predicted values are written as: $\hat y_i = \hat \beta_0 + \hat \beta_1 x_i$
- In other words, plug the value of the predictor variable $x$ into the linear equation
- The error or **residual** is $\hat e_i = y_i - \hat y_i$
- We'll go into more detail about residuals in a future lecture, but for now let's see an example

- What's the predicted life expectancy for the United States?

```{r}
gm2007 %>% 
  filter(country == "United States") %>%
  select(gdpPercap, lifeExp)
```

```{r}
59.6 + 0.000637 * 42951
```

- What's the residual?

```{r}
78.24 - 86.96
```


- Plotting all the predicted values

```{r}
gm2007$lifeExpPrediction <- predict(gm_model)
ggplot(gm2007, aes(gdpPercap, lifeExpPrediction)) +
  geom_point() + theme_tufte()
```

- Plotting all the residuals

```{r}
gm2007$errors <- residuals(gm_model)
qplot(gm2007$errors, bins = 20) + theme_tufte()
ggplot(gm2007, aes(gdpPercap, errors)) +
  geom_point() + 
  geom_hline(yintercept = 0) +
  theme_tufte()
```

- We'll come back to plots of residuals when we discuss regression diagnostics

### Extrapolation

- Can use the regression line to get predictions for new points, ones that are not in our original data
- A new point may not even have the outcome variable, but we can predict it using just the predictor variable
- But beware! It may not work as well...

- Suppose the population of Wakanda is about 12 million
- GDP: over 90 trillion (an estimate of T'Challa's net worth)
- GDP per capita is then at least 7,500,000
- Predicted life expectancy based on this calculation...

```{r}
59.6 + 0.000637 * (7500000)
```
- It can be dangerous to predict for values **outside the range** of the predictor in the data
- This is called **extrapolation** and it will probably yield larger errors than when predicting on values inside the original data

- Safer, and useful: predicting an outcome for a new point when the predictor variable lies **within the range** seen in the original data. (e.g. for `gdpPercap` between 277 and 49,357)
- Useful because we may not even have the actual outcome variable for a new data point. In this case, we can't calculate a residual

### Recap

- Plugging any value of $x$ into $\hat \beta_0 + \hat \beta_1 x$ gives a predicted outcome, but this may be a bad idea if $x$ is outside the range of the data
- Remember that the fitted regression line comes from an equation using **estimates** $\hat \beta_0, \hat \beta_1$

- A few things to think about:
- What would happen if we gathered a new sample of data and calculated the estimated coefficients again?
- How accurate are the estimates? Does the accuracy depend on sample size?
