---
title: "CILab"
author: "Your name here!"
date: "3/29/2018"
output: html_document
---
---
title: "CILab"
author: "Your name here!"
date: "3/29/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nycflights13)
library(fivethirtyeight)
```

## Calculate a confidence interval manually and plot it

Use the following data from a published study: means are 51.9 and 40.6, sample SDs are 23.3 and 25, and sample sizes are 44 and 58, respectively.

```{r}
Xbar1 <- 51.9
S1 <- 23.3
n1 <- 44
c1 <- qt(.975, n1 - 1)
err1 <- c1*S1/sqrt(n1)
lower1 <- Xbar1 - err1
upper1 <- Xbar1 + err1
c(lower1, upper1)
```

Copy and paste the above code. Change **all** variable names from var1 to var2 (double check!). Finally, change the numbers based on the information given above: mean 40.6, sample SD 25, and sample size 58.

# Delete this line and put your answer here

Next we combine the results into one data frame for plotting.

```{r}
data <- data.frame(sex = c("male", "female"),
                   lower = c(lower1, lower2),
                   upper = c(upper1, upper2))
ggplot(data) + 
  geom_errorbar(aes(x = sex, ymin = lower, ymax = upper)) +
  theme_minimal()
```

How would you interpret these results? 

# Delete this line and put your answer here

## How much money do movies make?

We'll use the `bechdel` dataset in the `fivethirtyeight` package. If you haven't already installed this package, you'll need to run `install.packages("fivethirtyeight")` first. In this data is a variable called `domgross` for domestic gross earnings of each film.

```{r}
# Remove movies that have NA values for some variables
movies <- bechdel[complete.cases(bechdel),]
# Histogram of earnings
qplot(movies$domgross, bins = 50)
```

Since this distribution is very skewed, the mean is not a very good summary statistic. Let's focus on the median instead. Fortunately, we can get a confidence interval for the median easily in `R` using the `wilcox.test` function.

```{r}
wilcox.test(movies$domgross, conf.int = TRUE)$conf.int
```
Is this different from the interval given by `t.test`? Input your code below to see, and explain your conclusion.

# Delete this line and put your answer here

Let's look at the return on investment by dividing by the `budget` variable.

```{r}
movies$return <- movies$domgross/movies$budget
```

- Create a histogram. Is the distribution very skewed? Does it make more sense to focus on the mean or median?
- Give a confidence interval for whichever parameter you decided to focus on.

# Delete this line and put your answer here

Now let's consider the returns separately based on whether movies pass or fail the Bechdel test (using the `binary` variable). To pass the Bechdel test, a movie has to meet 3 incredibly simple rules (and it is astonishing that there are so many movies that fail it).

```{r}
table(movies$binary)
```

1. It has to have at least two [named] women in it
2. Who talk to each other
3. About something besides a man

```{r}
movies %>% group_by(binary) %>% 
  summarize(median = median(return),
            mean = mean(return))
```

Based on this information, do you think it's safe to say that movies that **pass** the Bechdel test tend to have higher financial returns than those that fail it?

# Delete this line and put your answer here

## dplyr example creating many intervals

Change all appearances of AA (American Airlines) below to WN (Southwest).

```{r}
AA <- flights %>% filter(carrier == "AA")
head(AA %>% group_by(dest) %>% summarize(count = n()))
```

```{r}
AAtimes <- AA %>% group_by(dest) %>% summarize(
  lower = t.test(air_time)$conf.int[1],
  upper = t.test(air_time)$conf.int[2]) %>%
  arrange(lower)
head(AAtimes)
```

```{r}
ggplot() + geom_errorbar(data = AAtimes,
                aes(x = reorder(dest, lower),
                    ymin = lower,
                    ymax = upper)) +
  coord_flip() + theme_minimal()
```