---
title: "CILab"
author: "Your name here!"
date: "4/03/2018"
output: html_document
---
---
title: "CILab"
author: "Your name here!"
date: "4/03/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nycflights13)
library(fivethirtyeight)
library(MASS)
```

## Testing a single mean

Recall our 95 percent confidence interval for the flight time from NYC to Hawaii:

```{r}
X <- flights %>% filter(carrier == "HA") %>% pull(air_time)
t.test(X)$conf.int
```

- Test the null hypothesis that the mean flight time is for two choices of mu: one inside the interval and one outside.

```{r}
# Inside the interval:
t.test(X, mu = 620.8874)
```
Is the p-value larger or smaller than 0.05?

```{r}
# Outside the interval
```

Is the p-value larger or small than 0.05?


## Testing a single proportion

Suppose a polling company has surveyed 1000 registered voters and 551 of them said they supported a given candidate. Use `binom.test` to test the null hypothesis that the population support for the candidate is 50 percent against the one-sided alternative that it is greater.

```{r}
# code here. hint: binom.test(..., alt = "greater")
```

Is the p-value less than 0.05? Would the test reject the null hypothesis if the significance level was 1 percent?

## Testing two means

### Paired

Paired measurements mean that the two variables are measuring different aspects of the same individual or observational unit. For example, in the `mpg` data the observational units are cars, and two mileage variables are measured: highway mpg and city mpg. 

```{r}
mpg2008 <- mpg %>% filter(year == "2008")
# Use t.test with paired = TRUE
```

## Testing two medians

```{r}
movies <- bechdel[complete.cases(bechdel),]
movies$return <- movies$intgross_2013/movies$budget_2013
movies %>% group_by(binary) %>% 
  summarize(return = median(return))
# Movies that pass seem to have higher median returns
# But are we sure?
# New R feature: The "formula" below, with ~, splits
# the groups automatically using the binary variable
wilcox.test(return ~ binary, data = movies)
```
Would it make sense to do a paired test in this scenario?


## Does it make a difference which airline you take?

Use `t.test` with the formula to test if the flight time is different for each carrier.

```{r}
to_hawaii <- flights %>% filter(dest == "HNL") 
```
Which carrier has a longer flight time? Is that result significant at the 1 percent level?

## Test for independence

Most useful for kinds of data that are not numerical, but categorical. For example, the `survey` data from the `MASS` package has data from students at an Australian university

```{r}
head(survey)
```

Are the variables `Sex` and `Smoke` independent? 

```{r}
counts <- table(survey$Smoke, survey$Sex)
counts
```

Use `?chisq.test` to read about how you can test for independence, then use that function below on this data.

```{r}
# Code here
```

Does the data have sufficient evidence to reject the null hypothesis that the variables are independent at the 5 percent level? The 10 percent level?