---
title: "Homework 3 - R assignment"
author: (YOUR NAME) - I agree to abide by the Stern Code of Conduct
output: pdf_document
---
---
title: "Homework 3 - R assignment"
author: (YOUR NAME) - I agree to abide by the Stern Code of Conduct
output: pdf_document
---

```{r include=FALSE}
library(ggplot2)
```

### Question R1

This question will guide you through a *simulation study* in `R` to understand the bias of a certain estimator.

```{r}
# This creates a function that calculates sample variance
# but using n in the denominator instead of n-1
# you don't need to change anything here
varn <- function(x) mean((x - mean(x))^2)
```

#### Part a.

The code below generates a sample of size n and calculates both the sample variance and the biased version of sample version (the one that has n in the denominator instead of n-1). Modify the code to change n from 10 to some other value between 10 and 50. 

```{r}
# Create a sample of data by rolling a 6-sided die n times
n <- 10
data <- sample(1:6, n, replace = TRUE)
data
# True variance
35/12
# Unbiased estimate of variance
var(data)
# Biased estimate of variance
varn(data)
```

#### Part b.
Repeat the previous experiment many times and find the expected value of each variance estimator.

```{r}
# Unbiased estimator
mean(replicate(10000, var(sample(1:6, n, replace = TRUE))))
# Biased estimator
# copy the previous line here and change var to varn
```

#### Part c.
After running the previous code, which estimator's average value in the simulation is closer to the true value (roughly 2.9167)? 

\ 

Delete this line and type your answer here.


#### Part d.
Now go back and change n to another, larger value, and rerun all of the code. Do you notice anything about the average of the biased estimator?

\ 

Delete this line and type your answer here.

#### Part e.
Copy the code from part (b) and paste it below here, then change the `mean` function to be `sd` instead.

```{r}
# put code in here
```

Delete this sentence and replace with your answer: which estimator has more variability, the biased or unbiased one?

### Question R2

According to a Marketplace/Edison survey in April of 2017, about 23.4\% of survey responders agreed with the statement "the economic system in the U.S. is fair to all Americans." In this question we'll use a Bernoulli probability model to analyze this number. Suppose that there were 1,000 survey respondents and 234 agreed with the above quotation. Define a Bernoulli random variable which is 1 if a person agrees and 0 otherwise. Assume the survey was done with independent sampling (with replacement), so these Bernoulli random variables are independent. Then the number of people in the sample of 1,000 who agree is a Binomial random variable.

- We have $X_i$ i.i.d Ber($p$) for $i = 1, \ldots, 1000$.
- Let $S_n = \sum_{i=1}^n X_i$, so $S_n$ is Bin($n, p$).

### a. 
Using the fact that $n \bar X_n = S_n$, how could you use the Binomial distribution to calculate $P(a \leq \bar X_n \leq b)$?  How would you use `pbinom` with the given values of $a, b, n$, and $p$?

```
pbinom(something involving a, b, n, p) - pbinom(something involving a, b, n, p)
```

### b. 
Instead of the Binomial distribution, how would we use the central limit theorem to calculate the same probabilities? Hint: your answer should use `pnorm` and involve $\sqrt{n}$ (and $a, b$, and $p$).

```
pnorm(etc) ...
```

### c. 
Now let $n = 1000$, $p = 0.234$, $b = 0.250, a = 0.239$ and compute the desired probability with both methods.

\ 

```{r}
# write code here
```
