---
title: "Homework 7"
author: I agree to abide by the Stern Code of Conduct - (NAME & NYU ID)
output: html_document
---
---
title: "Homework 7"
author: I agree to abide by the Stern Code of Conduct - (NAME & NYU ID)
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# if you haven't installed ggfortify yet run this next line once without the "#" then delete the whole line
# install.packages("ggfortify") 
library(tidyverse)
library(ggfortify)
library(gapminder)
theme_set(theme_minimal())
options(scipen=999)
options(digits=4)
```

## Predicting life expectancy from GDP in the gapminder data

We will use data from the most recent year:

```{r}
gm2007 <- gapminder %>% filter(year == "2007")
```

In the rest of this homework, use the `gm2007` data to answer questions.

### Show first few rows of data below

```{r}

```

### Fit a linear model predicting life expectancy from GDP per capita

```{r}

```

### Summary of the model

```{r}

```

Do we reject the null hypothesis for the coefficient of GDP per capita at the 1\% level?

```
Answer
```

What proportion of the variance in life expectancy can be explained by this model?

```
Answer
```


### Confidence intervals for coefficients

```{r}

```

What is the lower limit of the 95\% confidence interval for the intercept?

```
Answer
```


### Diagnostic plots

Use the `autoplot()` function to show diagnostics below.

```{r}

```

Is there a pattern in the residuals? Do the residuals seem approximately normally distributed?

```
Answer
```

### Transforming to log(GDP per capita)

- Fit a model with `log10(gdpPercap)` as the predictor variable instead of `gdpPercap`
- Compute its summary, compare the value of R-squared to the one for the previous model


```{r}

```

Does the model (1) with `gdpPercap` or model (2) with `log10(gdpPercap)` do better at explaining the variance in life expectancy?

```
Answer
```

The coefficient for `log10(gdpPercap)` shows the increase in predicted life expectancy if gdpPercap increases by...

```
Answer
```

- Show diagnostic plots for this model.

```{r}

```

Do the residuals have any more or less of an obvious pattern in them? Do they seem any more or less normally distributed?

```
Answer
```

### If you could add one more variable, what would it be?

Suppose we could add another variable to this dataset to include in the model. It could be anything related to a country's economy, environment, consumption of specific goods, percent of population with access to high speed internet--anything like this that you can imagine. The goal is to better predict life expectancy and show that GDP and this other variable are important variables for predicting life expectancy.

Keep in mind that your ultimate goal is to tell a useful or interesting story with the data.

**(1)** Which variable would you pick, and why?

```
Answer
```

**(2)** How would you interpret the fitted coefficient for this variable after it is added to the model?

```
Answer
```

**(3)** Give an example of a variable which you think has a positive correlation with life expectancy, but a negative coefficient when included in a linear model with GDP as predictors. Explain your reasoning.

```
Answer
```

